---
title: "analysis answer"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
```

Load the data...

```{r}
temperature <- read_table(
    "https://chryswoods.com/data_analysis_r/cetml1659on.txt",
    skip = 6,
    na = c("-99.99", "-99.9"),
    col_types = cols("DATE" = col_integer())
)
```

Create the month levels

```{r}
month_levels <- c("JAN", "FEB", "MAR", "APR", "MAY", "JUN",
                  "JUL", "AUG", "SEP", "OCT", "NOV", "DEC")
```

Tidy the data...

```{r}
historical_temperature <- temperature %>%
    select(-YEAR) %>%
    pivot_longer(c("JAN", "FEB", "MAR", "APR", "MAY", "JUN",
                   "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"),
                 names_to = "month",
                 values_to = "temperature") %>%
    rename(year = DATE) %>%
    mutate(month = factor(month, month_levels))
```

Calculate the mean monthly temperatures in the 17th Century

```{r}
c17th <- historical_temperature %>%
    filter(year < 1700 & year >= 1600) %>%
    group_by(month) %>%
    summarise("temperature" = mean(temperature, na.rm = TRUE))
```

Calculate the mean monthly temperatures in the 21st Century

```{r}
c21st <- historical_temperature %>%
    filter(year >= 2000) %>%
    group_by(month) %>%
    summarise("temperature" = mean(temperature, na.rm = TRUE))
```

Now add the difference to the c21st table and print it out

```{r}
c21st <- c21st %>% 
    mutate(change = temperature - (c17th %>% select(temperature) %>% pull()))
c21st
```

From this we can see that most of the warming is focused on the winter
months.

We will now repeat this for the maximum and minimum temperatures...

```{r}
c17th_max <- historical_temperature %>%
    filter(year < 1700 & year >= 1600) %>%
    group_by(month) %>%
    summarise("temperature" = max(temperature, na.rm = TRUE))
c21st_max <- historical_temperature %>%
    filter(year >= 2000) %>%
    group_by(month) %>%
    summarise("temperature" = max(temperature, na.rm = TRUE))
c21st_max <- c21st_max %>% 
    mutate(change = temperature - (c17th_max %>% select(temperature) %>% pull()))
c21st_max
```

```{r}
c17th_min <- historical_temperature %>%
    filter(year < 1700 & year >= 1600) %>%
    group_by(month) %>%
    summarise("temperature" = min(temperature, na.rm = TRUE))
c21st_min <- historical_temperature %>%
    filter(year >= 2000) %>%
    group_by(month) %>%
    summarise("temperature" = min(temperature, na.rm = TRUE))
c21st_min <- c21st_min %>% 
    mutate(change = temperature - (c17th_min %>% select(temperature) %>% pull()))
c21st_min
```

Finally, we can get the average increase in monthly temperatures by calculating
the mean of the `change` column in `c21st`

```{r}
c21st %>% 
    select(change) %>% 
    pull() %>%
    mean()
```
